---
title: "class12"
author: "Tanieng Huang"
format: pdf
editor: visual
---

## Class12

section1-9

Package prep

```{r}
#install.packages("BiocManager")
#BiocManager::install()
# For this class we will need DESeq2:
#BiocManager::install("DESeq2")
library(BiocManager)
library(DESeq2)
```

### **3. Import countData and colData**

```{r}
# Complete the missing code
# Read in the metadata
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")

# Inspect the first few rows of each
head(counts)
head(metadata)
```

-   **Q1.** How many genes are in this dataset?

```{r}
nrow(counts)
```

> There are 38694 genes.

-   **Q2.** How many ‘control’ cell lines do we have?

```{r}
table(metadata$dex)
```

> 4 control cell lines.

### **4. Toy differential gene expression**

calculate the mean counts per gene across these samples:

```{r}
# Extract metadata rows where dex == "control"
control <- metadata[metadata[,"dex"]=="control",]

# Subset count data to only the control sample columns
control.counts <- counts[ ,control$id]

# Calculate the mean counts per gene across the 4 control samples
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

dplyr way

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

-   **Q3.** How would you make the above code in either approach more robust? Is there a function that could help here?

> Instead of hard-coding /4, use rowMeans() or divide by ncol(control.counts). If the number of control samples changes, the code still works correctly.

-   **Q4.** Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

```{r}
# Extract treated sample metadata
treated <- metadata[metadata[,"dex"]=="treated",]

# Subset counts to treated samples
treated.counts <- counts[, treated$id]

# Calculate mean per gene (using rowMeans for robustness)
treated.mean <- rowMeans(treated.counts)
head(treated.mean)
```

combine into a new frame:

```{r}
meancounts <- data.frame(control.mean, treated.mean)

# Check total counts per group
colSums(meancounts)
# Should show: control.mean = 23005324, treated.mean = 22196524
```

-   **Q5 (a).** Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts[,1], meancounts[,2], 
     xlab="Control", ylab="Treated")
```

**Q5 (b).**You could also use the **ggplot2** package to make this figure producing the plot below. What **geom\_?()** function would you use for this plot? 

> geom_point will be used.

```{r}
library(ggplot2)
ggplot(meancounts) + 
  aes(control.mean, treated.mean) + 
  geom_point(alpha=0.2)
```

**Q6.** Try plotting both axes on a log scale. What is the argument to **plot()** that allows you to do this? 

> log="xy" sets both axes to log scale in base R's plot() function.

```{r}
plot(meancounts[,1], meancounts[,2], 
     log="xy",
     xlab="log Control", ylab="log Treated")
```

calculate log2 fold change:

```{r}
# Add log2 fold change column: log2(treated/control)
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

Remove genes with zero counts (they produce NaN and -Inf):

```{r}
# Find rows where either control or treated mean is zero
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

# Get unique row indices to remove
to.rm <- unique(zero.vals[,1])

# Remove those rows
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

-   **Q7.** What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

> arr.ind=TRUE makes which() return both row and column indices as a matrix instead of just a single vector index. We take the first column \[,1\] because that gives us the row numbers of genes with zeros. We call unique() because a gene might have zero in both control AND treated columns. This would appear as two rows in the output, but we only want to remove that gene once.

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

-   **Q8.** Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

```{r}
sum(up.ind)
```

> 250 up regulated genes.

-   **Q9.** Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

```{r}
sum(down.ind)
```

> 367 down regulated genes.

-   **Q10.** Do you trust these results? Why or why not?

> No, we should not trust these results. We have no statistical test like p-values to determine if these differences are statistically significant. The fold change alone doesn't account for variability between replicates. A gene could show a large fold change due to random variation in one sample. We need a proper statistical framework like DESeq2 that accounts for both biological variability and sequencing depth differences.

### **5. Setting up for DESeq**

```{r}
library(DESeq2)

# Create the DESeqDataSet object
# countData = our count matrix
# colData = our metadata
# design = the experimental variable we care about (~dex)
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

### **6. Principal Component Analysis (PCA)**

```{r}
# Apply variance stabilizing transformation for visualization
vsd <- vst(dds, blind = FALSE)

# Plot PCA colored by treatment group
plotPCA(vsd, intgroup = c("dex"))
```

custom ggplot version:

```{r}
# Get PCA data
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)

# Calculate percent variance explained
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Custom ggplot PCA
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

### **7. DESeq analysis**

```{r}
# Run the full DESeq2 pipeline
dds <- DESeq(dds)

# Get results
res <- results(dds)
res
```

```{r}
summary(res)
```

stricter cutoff

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

### **8. Adding annotation data**

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
# See available key types
columns(org.Hs.eg.db)

# Add gene symbol column
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),       # Our Ensembl gene IDs
                     keytype="ENSEMBL",          # Format of our keys
                     column="SYMBOL",            # What we want to add
                     multiVals="first")          # If multiple matches, take first
```

```{r}
head(res)
```

-   **Q11.** Run the **mapIds()** function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     keytype="ENSEMBL",
                     column="ENTREZID",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                      keys=row.names(res),
                      keytype="ENSEMBL",
                      column="UNIPROT",
                      multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                       keys=row.names(res),
                       keytype="ENSEMBL",
                       column="GENENAME",
                       multiVals="first")

# View results ordered by adjusted p-value
ord <- order(res$padj)
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

### **9. Data Visualization**

#### **Volcano plots**

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

Add cutoff lines:

with the `abline()` function) and color (with a custom color vector) highlighting genes that have padj\<0.05 and the absolute log2FoldChange\>2.

```{r}
plot(res$log2FoldChange, -log(res$padj), 
     ylab="-Log(P-value)", xlab="Log2(FoldChange)")
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

Color-coded volcano plot:

```{r}
# Start with all points gray
mycols <- rep("gray", nrow(res))

# Color red if |log2FC| > 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue if ALSO padj < 0.01
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[ inds ] <- "blue"

# Plot with colors
plot(res$log2FoldChange, -log(res$padj), 
     col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)")
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
#| fig-width: 10
#| fig-height: 8
library(EnhancedVolcano)
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```
